Effects of epidemiological structure on the transient evolution of HIV virulence

### 2017-01-12 ms/ms.tex
fixing progresion time to Lspvl (and trying to merge)

\end{flushleft}
% Please keep the abstract below 300 words
% current is 296
\section*{Abstract}
Early in an epidemic, high densities of susceptible hosts
select for relatively high parasite virulence; later in the epidemic,
lower susceptible densities select for lower virulence.
Thus over the course of a typical epidemic the average virulence 
of parasite strains increases initially,
peaks partway through the epidemic, then declines again.
However, precise quantitative outcomes, such as the peak virulence reached
and its timing, may depend sensitively on epidemiological details. 
Fraser \emph{et al.}
proposed a model for the eco-evolutionary
dynamics of HIV that incorporates
the tradeoffs between transmission and virulence (mediated by
set-point viral load, SPVL) and their heritability between
hosts. Their models use implicit equations to
capture the effects of partnership dynamics that are at the core of 
epidemics of sexually transmitted diseases. 
Our models combine HIV virulence tradeoffs with a range of
contact models, modeling partnership formation and
dissolution and allowing for individuals to transmit disease outside
of partnerships. We assess summary statistics such as the peak virulence
(corresponding to the minimum expected time of progression to AIDS) across
models for a range of 
parameters applicable to the HIV epidemic in sub-Saharan Africa.
Although virulence trajectories are broadly similar
across models, the timing and magnitude of the 
virulence peak vary
considerably.
Previously developed implicit models predicted 
lower virulence and slower progression
at the peak (a minimum of 15 years to
progress to AIDS) compared both to more realistic models
and to simple random-mixing models with no partnership structure
at all (both with a minimum of $\approx$ 7.25 years to progress to AIDS).
In this range of models, the simplest random-mixing structure best
approximates the most realistic model; this
surprising outcome occurs because the dominance of extra-pair
contact in the realistic model swamps the effects of
partnership structure.

@@

Pathogens such as HIV can evolve rapidly when the environment changes.
Such changes include decreases in the probability that an infectious
contact (such as a sneeze for a respiratory disease, or an unprotected
sex act for a sexually transmitted disease) encounters an uninfected,
susceptible person and thus has a chance to transmit the
pathogen. Disease spread during an epidemic reduces the fraction of
susceptible people in the population and thus changes the direction of
parasite evolution.  While researchers have explored these
evolutionary dynamics with simulations, their models have neglected
important processes such as people entering and leaving sexual
partnerships. We compared several evolutionary models that include
partnership dynamics and sexual contact outside of stable
partnerships. Models of intermediate complexity predicted lower
virulence over the course of the epidemic (a minimum of 15 years to
progress to AIDS) than either more realistic models or simple models
without any partnership structure at all (both approx. 7.25 years to
progress to AIDS), because sexual contact outside of stable
partnerships tended to wash out the effects of contact
structure. The large effect of epidemiological details on evolutionary
dynamics suggests that researchers trying to predict the evolution of
pathogens should proceed with caution.

@@

As in Shirreff \etal\ \cite{shirreff_transmission_2011}, our models
explicitly track the evolution of mean $\log_{10}$ SPVL
(which we denote as $\alpha$), rather than the rate of progression to
AIDS itself (hereafter ``virulence'' will refer either to the SPVL, or
to the rate of progression to AIDS; these two quantities are
deterministically linked in the model).  
In contrast to Shirreff \etal, we use a
single-stage disease model instead of accounting explicitly for
progression through the three main stages of HIV infection (primary,
asymptomatic, and disease), and we use an exponentially
distributed infectious period instead of a more realistic
Weibull-distributed infectious period; we show below that our results
are not overly sensitive to this simplification. We account for
varying transmission rates and durations of each disease stage by
summing the durations of three stages (again based on Shirreff \etal's
model) and taking the duration-weighted average of transmission rates
of three stages. Thus the within-couple transmission rate, $\beta$,
for our models is given by:
\begin{equation}
\beta (\alpha) = \frac{D_P \beta_P + D_A (\alpha) \beta_A (\alpha) + D_D \beta_D}{D_P + D_A (\alpha) + D_D},
\end{equation}
where the duration of infection ($D_P$ and $D_D$) and rate of transmission ($\beta_P$ and $\beta_D$) of the Primary and Disease stages
of infection are independent of the host's SPVL
(Table~\ref{table:parmsTable} gives definitions, units, and values for all parameters).
. Following Shirreff \etal, the duration of infection ($D_A$) and rate of transmission ($\beta_A$) for the Asymptomatic stage are Hill functions of the SPVL:

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../figures/fig1.pdf}
\caption{{\bf Baseline dynamics.}
Time series of mean population \Lspvl\ using baseline parameters. Each initial condition was varied independently in order to test their effects on transient dynamics of virulence evolution. Note that the initial condition has negligible effect on equilibrium virulence.
(a) Contrast between the three-stage Shirreff model and the single-stage model calibrated to varying initial exponential growth rates, $r$.
(b) Effects of varying initial infectious density $I(0)$.
(c) Effects of varying initial mean virulence $\alpha(0)$.
The middle $r=0.042$ (orange, dotted) curve in panel (a), calibrated to match the epidemic dynamics of Shirreff \etal's three-stage model \cite{shirreff_transmission_2011}, shows that our simplified single-stage HIV model can produce similar SPVL trajectories to the original three-stage model; the other $r$ values are chosen to show the effects of doubling or halving the initial epidemic growth rate. 
The eco-evolutionary dynamics are qualitatively similar
across a range of initial conditions (panels b,c).}
\label{fig:panel3}
\end{figure}

@@

\subsection*{Initial conditions}

Choosing the initial conditions for the simulations is something of a
challenge. In the most common theoretical modeling contexts
researchers are primarily interested in equilibria (or other long-term
dynamical attractors such as limit cycles) and are exploring models
that have a single stable attractor, so initial conditions can be
ignored as long as we run models for long enough.  In eco-evolutionary
dynamics, however, the initial conditions do affect our
conclusions. We have no empirical information that would justify a
particular quantitative choice of the fraction infected and the mean
and variance of the distribution of SPVL at the point when the
pandemic strain of HIV-1 entered the human population; in any case,
the level of realism of our model does not support such a detailed
consideration of the early dynamics of HIV.  In most cases, we started
with an initial \Lspvl\ of 3.0, to match the value used by
Shirreff \etal. Shirreff \etal\ use an initial $I(0)=10^{-3}$;
we show results for this value in \figurename~\ref{fig:panel3}.
Since some parameters in our sampling process (see next section)
lead to very low equilibrium prevalence, we switched to 
$I(0)=10^{-4}$ for the remainder of our analysis.
We ran one small set of sensitivity analyses with baseline values of the
epidemiological parameters (Table~1) to test the effects of the
initial conditions. Appendix~S1 provides
further details on initial conditions, such
as the initial distribution of SPVL around the mean and the
distribution of initial infected density across single people
and different partnership types.

@@

Our simplifications of Shirreff \etal's model \cite{shirreff_transmission_2011} reproduce its qualitative behaviour --- in particular, its predictions of virulence dynamics --- reasonably well. As we change $r$ from 0.42 to 0.84 (the former value matching the initial rate of increase in prevalence in Shirreff \etal's full model) the initial trajectory of increasing virulence brackets the rate from the original model (\figurename~\ref{fig:panel3}a). However, our model produces lower peak virulence ($\approx 4.3$ vs. $\approx 4.6$ \Lspvl) 
and equilibrium virulence ($\approx 4.25$ vs. $\approx 4.5$ \Lspvl) than Shirreff's, even for matching initial incidence trajectories (i.e., $r=0.042~\textrm{year}^{-1}$).

Changing the initial infectious density ($I(0)$), while it produces the expected changes in the initial epidemic trajectory (Supplementary material), has little effect on the virulence trajectory. Decreasing $I(0)$ makes SPVL peak slightly later and higher, because it allows a longer epidemic phase before the transition to endemic dynamics (\figurename~\ref{fig:panel3}b). Decreasing the initial SPVL also leads to progressively later, higher peaks in SPVL (\figurename~\ref{fig:panel3}c). In this case the delay in the peak is more pronounced than for low $I(0)$, because the rate of SPVL increase is eventually limited by the mutation rate. The peak SPVL is actually larger for a lower starting SPVL, presumably because lower SPVL also allows for a longer epidemic phase. However, the peaks are similar across the entire range of initial conditions, because even in the most limited (high-$I(0)$, high-$\alpha(0)$) cases HIV is able to evolve most of the way to the SPVL that would be optimal in an indefinitely prolonged growth phase.

>>>>

\end{flushleft}
% Please keep the abstract below 300 words
% current is 296
\section*{Abstract}
Early in an epidemic, high densities of susceptible hosts
select for relatively high parasite virulence; later in the epidemic,
lower susceptible densities select for lower virulence.
Thus over the course of a typical epidemic the average virulence 
of parasite strains increases initially,
peaks partway through the epidemic, then declines again.
However, precise quantitative outcomes, such as the peak virulence reached
and its timing, may depend sensitively on epidemiological details. 
Fraser \emph{et al.}
proposed a model for the eco-evolutionary
dynamics of HIV that incorporates
the tradeoffs between transmission and virulence (mediated by
set-point viral load, SPVL) and their heritability between
hosts. Their models use implicit equations to
capture the effects of partnership dynamics that are at the core of 
epidemics of sexually transmitted diseases. 
Our models combine HIV virulence tradeoffs with a range of
contact models, modeling partnership formation and
dissolution and allowing for individuals to transmit disease outside
of partnerships. We assess summary statistics such as the peak virulence
(corresponding to the maximum value of population mean \Lspvl\ achieved throughout the epidemic) across
models for a range of 
parameters applicable to the HIV epidemic in sub-Saharan Africa.
Although virulence trajectories are broadly similar
across models, the timing and magnitude of the 
virulence peak vary
considerably.
Previously developed implicit models predicted 
lower virulence and slower progression
at the peak (a maximum of 3.5 \Lspvl) compared both to more realistic models
and to simple random-mixing models with no partnership structure
at all (both with a maximum of $\approx$ 4.7 years to \Lspvl).
In this range of models, the simplest random-mixing structure best
approximates the most realistic model; this
surprising outcome occurs because the dominance of extra-pair
contact in the realistic model swamps the effects of
partnership structure.

@@

Pathogens such as HIV can evolve rapidly when the environment changes.
Such changes include decreases in the probability that an infectious
contact (such as a sneeze for a respiratory disease, or an unprotected
sex act for a sexually transmitted disease) encounters an uninfected,
susceptible person and thus has a chance to transmit the
pathogen. Disease spread during an epidemic reduces the fraction of
susceptible people in the population and thus changes the direction of
parasite evolution.  While researchers have explored these
evolutionary dynamics with simulations, their models have neglected
important processes such as people entering and leaving sexual
partnerships. We compared several evolutionary models that include
partnership dynamics and sexual contact outside of stable
partnerships. Models of intermediate complexity predicted lower
virulence over the course of the epidemic (a maximum of 3.5 \Lspvl) than either more realistic models or simple models
without any partnership structure at all (both approx. 4.7 \Lspvl), because sexual contact outside of stable
partnerships tended to wash out the effects of contact
structure. The large effect of epidemiological details on evolutionary
dynamics suggests that researchers trying to predict the evolution of
pathogens should proceed with caution.

@@

As in Shirreff \etal\ \cite{shirreff_transmission_2011}, our models
explicitly track the evolution of mean $\log_{10}$ SPVL
(which we denote as $\alpha$), rather than the rate of progression to
AIDS itself (hereafter ``virulence'' will refer to \Lspvl).  
In contrast to Shirreff \etal, we use a
single-stage disease model instead of accounting explicitly for
progression through the three main stages of HIV infection (primary,
asymptomatic, and disease), and we use an exponentially
distributed infectious period instead of a more realistic
Weibull-distributed infectious period; we show below that our results
are not overly sensitive to this simplification. We account for
varying transmission rates and durations of each disease stage by
summing the durations of three stages (again based on Shirreff \etal's
model) and taking the duration-weighted average of transmission rates
of three stages. Thus the within-couple transmission rate, $\beta$,
for our models is given by:
\begin{equation}
\beta (\alpha) = \frac{D_P \beta_P + D_A (\alpha) \beta_A (\alpha) + D_D \beta_D}{D_P + D_A (\alpha) + D_D},
\end{equation}
where the duration of infection ($D_P$ and $D_D$) and rate of transmission ($\beta_P$ and $\beta_D$) of the Primary and Disease stages
of infection are independent of the host's SPVL
(Table~\ref{table:parmsTable} gives definitions, units, and values for all parameters).
. Following Shirreff \etal, the duration of infection ($D_A$) and rate of transmission ($\beta_A$) for the Asymptomatic stage are Hill functions of the SPVL:

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../figures/fig1.pdf}
\caption{{\bf Baseline dynamics.}
Time series of mean population \Lspvl\ using baseline parameters. Each initial condition was varied independently in order to test their effects on transient dynamics of virulence evolution. Note that the initial condition has negligible effect on equilibrium virulence.
(a) Contrast between the three-stage Shirreff model and the single-stage model calibrated to varying initial exponential growth rates, $r$. achieved quicker.
(b) Effects of varying initial infectious density $I(0)$.
(c) Effects of varying initial mean virulence $\alpha(0)$.
The $r=0.042$ (orange, dotted) curve in panel (a), calibrated to match the epidemic dynamics of Shirreff \etal's model \cite{shirreff_transmission_2011}, shows that our simplified model can produce similar virulence trajectories. 
Although virulence dynamics are sensitive to initial conditions, we are still able to achieve similar qualitative dynamics dynamics across a range of parameters. 
For fair comparison, we hold them constant throughout our simulations.}
\label{fig:panel3}
\end{figure}

@@

@@

Our simplifications of Shirreff \etal's model \cite{shirreff_transmission_2011} reproduce its qualitative behaviour --- in particular, its predictions of virulence dynamics --- reasonably well. As $r$ decreases from 0.084 to 0.42 (the latter value matching the initial rate of increase in prevalence in Shirreff \etal's full model) the initial trajectory of increasing virulence brackets the rate from the original model (\figurename~\ref{fig:panel3}a). However, our model produces lower peak virulence ($\approx 4.3$ vs. $\approx 4.6$ \Lspvl) 
and equilibrium virulence ($\approx 4.25$ vs. $\approx 4.5$ \Lspvl) than Shirreff's, even for matching initial incidence trajectories (i.e., $r=0.042~\textrm{year}^{-1}$).

Changing the initial infectious density ($I(0)$), while it produces the expected changes in the initial epidemic trajectory (Supplementary material), has little effect on the virulence trajectory, making the virulence peaks slightly later and larger as $I(0)$ decreases. Decreasing $I(0)$ allows a longer epidemic phase before the transition to endemic dynamics (\figurename~\ref{fig:panel3}b). Decreasing the initial virulence
also leads to progressively later, larger peaks in virulence (\figurename~\ref{fig:panel3}c).
